Semantic Novelty Detection with Graphical Models

### 2011-05-05 semantic-nd/article.tex
Added drawings and refactor novelty part.

% Explain approach of using conditional probability
\subsection{Conditional Probability: $P(x|\overline{novel})$}
One of the hardest problems is that on most cases novelty detection system only have access to 
the known data.
Under that the best approach is to define a threshold based on $P(x|\overline{novel})/P(x)$
considering that $P(x)$ is constant through all the input space.

In the presented case, $P(x|\overline{novel})$ is approximated by the graphical model $G$ produced
by the conceptual layer. As so we use $P_G(x)$ as a substitute.

\label{sec:uniform-unconditional}
In several cases assuming the unconditional probability $P(x)$ is constant leads to discarding the factor,
nonetheless here the distributions are dynamically changing as the robot learns more on the environment.
So we need to keep the normalizing argument $P(x)$ and re-calculate it for each new subset of $x$.
Still assuming that the unconditional distributions generates all possible outcome with the same probability
we can model it $P(x)=\prod{|x_i|}$, where $|x_i|$ denotes the state space size of variable $x_i$.
In graphical model terms, this would be equivalent to build a factor graph $G'$ with the same variables but
without any factors.

Using a graphical model built to model the known data and assuming that in the unconditional case all possible outcomes
are equally likely a novelty threshold would be given by: $P_G(x)/P_{G'}(x)$, as illustrated on \autoref{fig:unsupervised-threshold}.

\begin{figure}[h]
\centering
\subfloat[Factor graph $G$ used to model conditional probability]{
\begin{tikzpicture}
  \matrix[row sep=1em,column sep=1.5em, ampersand replacement=\&] {
    \& \node [var] (room) {Room Category}; \& \\

    \node [factor] (fAppearance) {}; \&
    \node [factor] (fShape) {}; \&
    \node [factor] (fSize) {}; \\

    \node [var] (appearance) {Appearance}; \&
    \node [var] (shape) {Shape}; \&
    \node [var] (size) {Size}; \\
  };
  \path [line] (room) -- (fAppearance);
  \path [line] (room) -- (fShape);
  \path [line] (room) -- (fSize);

  \path [line] (fAppearance) -- (appearance);
  \path [line] (fShape) -- (shape);
  \path [line] (fSize) -- (size);
\end{tikzpicture}}
\hfill
\subfloat[Factor graph $G'$ used to model a unconditional probability distribution]{
\begin{tikzpicture}
  \matrix[row sep=1em,column sep=1.5em, ampersand replacement=\&] {
    \& \node [var] (room) {Room Category}; \& \\

    \node [var] (appearance) {Appearance}; \&
    \node [var] (shape) {Shape}; \&
    \node [var] (size) {Size}; \\
  };
\end{tikzpicture}}

\caption{\label{fig:unsupervised-threshold}
         Assuming a uniform unconditional distribution of any subset of the variables $x$,
         a novelty threshold can be defined on the function $P_G(x)/P_{G'}(x)$.}
\end{figure}

\subsection{Using Unlabelled Data}
Nonetheless in several cases its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability distribution that
the uniform one assumed in \autoref{sec:uniform-unconditional}.

For interesting and practical reasons it was assumed that the unconditional distribution could be modelled with
all the variables independent from each other.
Which is the same as assuming $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate $P_{x_i}$.
Once again \autoref{fig:semisupervised-threshold} illustrates this in terms of graphs $G$ and $G'$.

\begin{figure}[h]
\centering
\subfloat[Factor graph $G$ used to model conditional probability]{
\begin{tikzpicture}
  \matrix[row sep=1em,column sep=1.5em, ampersand replacement=\&] {
    \& \node [var] (room) {Room Category}; \& \\

    \node [factor] (fAppearance) {}; \&
    \node [factor] (fShape) {}; \&
    \node [factor] (fSize) {}; \\

    \node [var] (appearance) {Appearance}; \&
    \node [var] (shape) {Shape}; \&
    \node [var] (size) {Size}; \\
  };
  \path [line] (room) -- (fAppearance);
  \path [line] (room) -- (fShape);
  \path [line] (room) -- (fSize);

  \path [line] (fAppearance) -- (appearance);
  \path [line] (fShape) -- (shape);
  \path [line] (fSize) -- (size);
\end{tikzpicture}}
\hfill
\subfloat[Factor graph $G'$ used to model a unconditional probability distribution]{
\begin{tikzpicture}
  \matrix[row sep=1em,column sep=1.5em, ampersand replacement=\&] {
    \node [var] (appearance) {Appearance}; \&
    \node [var] (shape) {Shape}; \&
    \node [var] (size) {Size}; \\

    \node [factor] (fAppearance) {}; \&
    \node [factor] (fShape) {}; \&
    \node [factor] (fSize) {}; \\
  };
  \path [line] (fAppearance) -- (appearance);
  \path [line] (fShape) -- (shape);
  \path [line] (fSize) -- (size);
\end{tikzpicture}}

\caption{\label{fig:semisupervised-threshold}Using unlabelled data, its possible to create a better approximation
         for the unconditional probability. Which should lead to a better novelty threshold.}
\end{figure}

The described ratio between the two graphs can be understood as compensating for an existing bias
on the unconditional distribution.
Which is an important step to achieve a correct ordering of the input space for novelty thresholding.
This way our unconditional function perceive which features are slightly more biased to a certain value and will
compensate on the probability calculation.

>>>>

% Explain approach of using conditional probability
\subsection{Conditional Probability: $P(x|\overline{novel})$}
\label{sec:uniform-unconditional}

One of the hardest problems is that on most cases a novelty detection system only have access to 
the known data.
Under that the best approach is to define a threshold based on $P(x|\overline{novel})/P(x)$
considering that $P(x)$ is constant through all the input space.

In the presented case the conditional probability is approximated by the graphical model $G$
produced by the conceptual layer.
The \autoref{fig:conditional-prob-graph} represents a possible graph $G$ built from the conceptual
layer to represent the conditional probability on the sensed variables $x$.
$P_G(x)$ is then used as an approximation for $P(x|\overline{novel})$.

\begin{figure}[h]
\centering
\includegraphics[width=0.70\textwidth]{figures/conditional-prob-graph.pdf}
\caption{\label{fig:conditional-prob-graph}The conceptual layer models the conditional
         probability distribution of the sensed variables by using hidden variables.
         Those are created from the conceptual knowledge and represent the structural
         and connectivity information the system is aware of.
         In this case only hidden variables $R_i$ were used to model the room categories
         where each property was sensed.}
\end{figure}

Its important to notice that in several cases assuming the unconditional probability $P(x)$ is constant leads to discarding the factor.
Nonetheless, here the distributions are dynamically changing as the system learns more on the environment.
So we need to keep the normalizing argument $P(x)$ and re-calculate it for each new subset of $x$.

Assuming that the unconditional distributions generates all possible outcome with the same probability
we can model it with $\prod{1/|x_i|}$, where $|x_i|$ denotes the state space size of variable $x_i$.
In graphical model terms, this would be equivalent to build a factor graph $U$ with the same variables but
without any factors as illustrated on \autoref{fig:uniform-prob-graph}.

\begin{figure}
\centering
\includegraphics[width=0.70\textwidth]{figures/uniform-prob-graph.pdf}
\caption{\label{fig:uniform-prob-graph}Without any existing factors, this graph $U$ represents a uniform
         distribution over the set $x$.}
\end{figure}

Having now a graphical model built to model the known data distribution and assuming that in the unconditional case all possible outcomes are equally likely a novelty threshold would be given by: $P_G(x)/P_{U}(x)$.

\subsection{Using Unlabelled Data}
Nonetheless in several cases its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability distribution than
the uniform one assumed in \autoref{sec:uniform-unconditional}.

For interesting and practical reasons it was assumed that the unconditional distribution
could be modelled with all the variables independent from each other.
Which translates as $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate $P_{x_i}$.
\autoref{fig:semisupervised-threshold} illustrates this in terms of graphs.

\begin{figure}[h]
\centering
\includegraphics[width=0.70\textwidth]{figures/independent-prob-graph.pdf}
\caption{\label{fig:semisupervised-threshold}Using unlabelled data, its possible to estimate
         the unconditional probability for each variable and this way try to compensate
         for a possible existent bias on some of the variables.}
\end{figure}

Once again the novelty threshold would be given by $P_G(x)/P_U(x)$. This ratio between the two graphs
can be understood as compensating for an existing bias on the unconditional distribution.
Which is an important step to achieve a correct ordering of the input space for novelty thresholding.
This way our unconditional function perceives which features are slightly more biased to a certain value
and will compensate on the novelty calculation.
